The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-13 talk/vmil2012/paper.tex
start adding David Edelsohn's comments

\hypersetup{%
  plainpages=false,%
  %hyperfootnotes=false,%
  colorlinks=true,%
  urlcolor=black,%
  citecolor=black,%
  linkcolor=black,%
  pdftitle={Efficiently Handling Guards in the Low Level Design of RPython's Tracing JIT},%
  pdfauthor={David Schneider},
}

@@

\title{Efficiently Handling Guards in the Low Level Design of RPython's Tracing JIT}

@@

In this paper we describe and analyze how deoptimization works in the context
of tracing just-in-time compilers. What instructions are used in the
intermediate and low-level representation of the JIT instructions and how these
are implemented.

Although there are several publications about tracing just-in-time compilers,
to our knowledge, there are none that describe deoptimization and the use and
implementation of guards in this context.

Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT and that guards are
operations that are associated with an overhead to maintain information about
the execution state to be able to rebuild it in case of deoptimization, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of RPython's
tracing JIT and explain the rationale behind the design decisions based on the
numbers provided here.

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed, this process is described in more detail in
Section~\ref{sec:Resume Data}, during the recording phase special operations,
referred to as \texttt{guards}, are inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for 14.42\% to 22.32\% of the
operations before and for 15.2\% to 20.12\% of the operations after the
optimization pass over the traced and later compiled parts of the benchmarks,
making guards one of the most common types of operations. Many of these guards
fail rarely or not all during execution. There are several aspects to consider
in the design and optimization of guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace,  - which is
the common case - should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the low-level design of a tracing just-in-time compiler.

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind and the implementation of guards in
RPython's tracing just-in-time compiler, the contributions of this paper are:
\begin{itemize}
  \item An analysis of guards in the context of RPython's tracing JIT to
  substantiate the aforementioned observation, based on a set of benchmarks.
  \item We provide a detailed measurements about the frequency and the
  overhead associated with guards.
  \item We provide a description about how guards are implemented in the high\-
  and low-level parts of the JIT and describe the rationale behind the design.
\end{itemize}
\begin{figure}
    \include{figures/guard_table}
    \caption{Percentage of guards before and after optimization for different benchmarks}
    \label{fig:guard_percent}
\end{figure}

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend\bivab{better term for this?} related to recording and storing the
information required to restore the interpreter state in case of a guard
failure, once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

The RPython language and the PyPy Project were started in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative implementation of the Python language\bivab{mention speed}. The
implementation takes advantage of the language features provided by RPython
such as the provided tracing just-in-time compiler described below.

@@

RPython is built of two components, the language and the translation toolchain
used to transform RPython programs to executable units.  The RPython language
is a statically typed object oriented high level language. The language provides
several features such as automatic memory management (aka. Garbage Collection)
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the program to a low level representations suited to be compiled
and run on one of the different supported target platforms/architectures such
as C, .NET and Java. During the transformation process
different low level aspects suited for the target environment are automatically
added to program such as (if needed) a garbage collector and with some hints
provided by the author a just-in-time compiler.

\subsection{RPython's Tracing JIT Compilers}
\label{sub:tracing}
Tracing JITs are a technique of just-in-time compilers that generate code by
observing the execution of a program. VMs using tracing JITs are typically
mixed mode execution environments containing also an interpreter. The
interpreter profiles the executed program and selects frequently executed code
paths to be compiled to machine code. After profiling identified an interesting
path, tracing is started, recording all operations that are executed on this
path. Like in most compilers tracing JITs use an intermediate representation to
store the recorded operations, which is typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}, these operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

@@

Since tracing linearizes control flow by following one concrete execution,
not the full control flow of a program is observed.
The possible points of deviation from the trace are guard operations
that check whether the same assumptions observed during tracing still hold during execution.
In later executions of the trace the guards can fail.
If that happens, execution needs to continue in the interpreter.
This means it is necessary to attach enough information to a guard
to reconstruct the interpreter state when that guard fails.
This information is called the \emph{resume data}.

@@

The core idea of storing resume data as compactly as possible
is to share parts of the data structure between subsequent guards.
This is often useful because the density of guards in traces is so high,
that quite often not much changes between them.
Since resume data is a linked list of symbolic frames
often only the information in the top frame changes from one guard to the next.
The other frames can often be just reused.
The reason for this is that during tracing only the variables
of the currently executing frame can change.
Therefore if two guards are generated from code in the same function
the resume data of the rest of the frame stack can be reused.

@@

The other important point of interaction between resume data and the optimizer
is RPython's allocation removal optimization~\cite{bolz_allocation_2011}.
This optimization discovers allocations in the trace that create objects
that do not survive long.
An example is the instance of \lstinline{Even} in the example\cfbolz{reference figure}.
Allocation removal makes resume data more complex.
Since allocations are removed from the trace it becomes necessary
to reconstruct the objects that were not allocated so far when a guard fails.
Therefore the resume data needs to store enough information
to make this reconstruction possible.

Adding this additional information is done as follows.
So far, every variable in the symbolic frames
contains a constant or an SSA variable.
After allocation removal the variables in the symbolic frames can also contain
``virtual'' objects.
These are objects that were not allocated so far,
because the optimizer removed their allocation.
The virtual objects in the symbolic frames describe exactly
how the heap objects look like which have to be allocated on guard failure.
To this end, the content of every field of the virtual object is described
in the same way that the local variables of symbolic frames are described.
The fields of the virtual objects can therefore be SSA variables, constants
or other virtual objects.
They are encoded using the same compact two-byte representation
as local variables.

@@

Figure~\ref{fig:trace-log} shows the optimized version of the trace in
Figure~\ref{fig:fig:unopt-trace}. Allocation removal has removed the
\lstinline{new} operation and other operations handling the boxes. The
operations handle unboxed numbers now.

@@

After optimization the resulting trace is handed to the over platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward one to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the to be
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics, operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can merged, reducing even more the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how an \texttt{int\_eq} operation
followed by a guard that checks the result of the operation are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

Each guard in the IR has attached to it a list of the IR-variables required to
rebuild the execution state in case the trace is left through the side-exit
corresponding to the guard. When a guard is compiled, additionally to the
condition check two things are generated/compiled. First a special data
structure called \emph{low-level resume data} is created that encodes the
information provided by the register allocator about where the values
corresponding to each IR-variable required by the guard will be stored when
execution reaches the code emitted for the corresponding guard. This data
structure stores the data in a compressed manner using an encoding the uses
8bits to store 7bits of information. This encoding is efficient to create and
provides a compact representation of the needed information. This encoding
needs to be as compact as possible to maintain an acceptable memory profile.

Second a piece of code is generated for each guard that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline. In case the
condition checked in the guard fails execution and a side-exit should be taken
execution jumps to the trampoline. In the trampoline the pointer to the
\emph{low-level resume data} is loaded and jumps to generic bail-out handler
that is used to leave the compiled trace in case of a guard failure.

Using the encoded location information the bail-out handler reads from the
saved execution state the values that the IR-variables had  at the time of the
guard failure and stores them in a location that can be read by the fronted.
After saving the information the control is passed to the frontend signaling
which guard failed so the frontend can read the information passed and restore
the state corresponding to the point in the program.

As in previous sections the underlying idea for the design of guards is to have
a fast on-trace profile and a potentially slow one in the bail-out case where
the execution takes one of the side exits due to a guard failure. At the same
time the data stored in the backend needed to rebuild the state needs to be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

As explained in previous sections, when a specific guard has failed often enough
a new trace, referred to as a \emph{bridge}, starting from this guard is recorded and
compiled. When compiling bridges the goal is that future failures of the guards
that led to the compilation of the bridge should execute the bridge without
additional overhead, in particular the failure of the guard should not lead
to leaving the compiled code prior to execution the code of the bridge.

The process of compiling a bridge is very similar to compiling a loop.
Instructions and guards are processed in the same way as described above. The
main difference is the setup phase. When compiling a trace we start with a clean
slate. The compilation of a bridge is started from a state (register and stack
bindings) that corresponds to the state during the compilation of the original
guard. To restore the state needed to compile the bridge we use the encoded
representation created for the guard to rebuild the bindings from IR-variables
to stack locations and registers used in the register allocator.  With this
reconstruction all bindings are restored to the state as they were in the
original loop up to the guard.

Once the bridge has been compiled the guard that led to compiling the bridge is
patched to redirect control flow to the bridge in case the check fails. In
future if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a digram of a compiled loop with two guards,
Guard \#1 jumps to the trampoline, loads the \texttt{low level resume data} and
then calls the compensation code, whereas Guard \#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

\section{Evaluation}
\label{sec:evaluation}
\todo{improve the table formatting}

@@

For the different benchmarks the \texttt{low-level resume data} has a size of
about 15\% to 20\% of the amount of memory compared to the size of the
generated machine code. On the other hand the generated machine code has only a
size ranging from 20.5\% to 37.98\% of the size of the high and low-level
\texttt{resume data} combined and being compressed as described before.

Tracing JIT compilers only compile the subset of the code executed in a program
that is traced in a hot loop, for this reason the amount of generated machine
code will be smaller than in other juts-in-time compilation approaches.  This
creates a larger discrepancy between the size of the \texttt{resume data} when
compared to the illustrates why it is important to compress this information.

@@

\subsection{Guard Failures}
\label{sub:guard_failure}
The last point in this discussion is the frequency of guard failures.
Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
relative amounts of guards that ever fail and of guards that fail more than 200
times.\footnote{
    The threshold of 200 is rather high. It was picked experimentally to give
    good results for long-running programs.
}
As described before, for guards that fail more than 200 times, a trace
is recorded that starts from the guard. Afterwards the guard is patched so that later
failures execute the new trace instead of taking the side-exit. Hence the
numbers presented for guards that fail more than 200 times represent the 200
failures up to the compilation of the bridge and all executions of the then
attached bridge.

@@

From Figure~\ref{fig:failing_guards} we can see that only a very small amount
of all the guards in the optimized traces ever fail. This amount varies between
2.4\% and 5.7\% of all guards. As can be expected, even less guards fail often
enough that a bride is compiled for them, only 1.2\% to 3.6\% of all guards
fail more than 200 times. Also of all failing guards a few fail extremely often
and most fail rarely. The results emphasizes that as most of the guards never
fail it is important to make sure that the successful execution of a guard does
not have unnecessary overhead.

@@

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information
about how to rebuild the state from a side-exit using the information in the
snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
snapshots for guards in LuaJIT are associated with a large memory footprint.
The solution used in there is to store sparse snapshots, avoiding the creation
of snapshots for every guard to reduce memory pressure. Snapshots are only
created for guards after updates to the global state, after control flow points
from the original program and for guards that are likely to fail. As an outlook
Pall mentions the plans to switch to compressed snapshots to further reduce
redundancy. The approach of not creating snapshots at all for every guard is
orthogonal to the resume data compression presented in this paper and could be
reused within RPython to improve the memory usage further.

@@

The deoptimization information of Hotspot is extended
to support correct behaviour
when scalar replacement of fields is done for non-escaping objects~\cite{kotzmann_escape_2005}.
The approach is extremely similar to how RPython's JIT handles virtual objects.
For every object that is not allocated in the code,
the deoptimization information contains a description
of the content of the fields.
When deoptimizing code, these objects are reallocated
and their fields filled with the values
described by the deoptimization information.
The paper does not describe any attempts to store this information compactly.
This may not be needed in their approach, because method-based JITs have a lot
fewer deoptimization points than tracing JITs.

@@

Finally we have presented experimental data collected using the standard PyPy
benchmark set to evaluate previous observations and assumptions. Our
experiments showed that, as previously assumed, guards are a very common
operation in traces. At the same time guards are associated with a high
overhead, because for all compiled guards information needs to be
stored to restore the execution state in case of a bail-out. The measurements
showed that the compression techniques used in PyPy effectively reduce the
overhead of guards, while it still produces a significant overhead. The results
also showed that guard failure is a local event: there are few
guards that fail at all, and even fewer that fail very often.
These numbers validate the design decision of reducing the overhead of
successful guard checks as much as possible while paying a higher price in the
case of bailout due to having to decode compressed state representation.
The compressed state representation is reduces the memory footprint of rarely
used data.

>>>>

\hypersetup{%
  plainpages=false,%
  %hyperfootnotes=false,%
  colorlinks=true,%
  urlcolor=black,%
  citecolor=black,%
  linkcolor=black,%
  pdftitle={The Efficient Handling of Guards in the Design of RPython's Tracing JIT},%
  pdfauthor={David Schneider},
}

@@

\title{The Efficient Handling of Guards in the Design of RPython's Tracing JIT}

@@

\todo{the introduction needs some work}
\cfbolz{the first two two paragraphs talk about deoptimization, then it
switches to guards. I would say we should only talk about guards in the
beginning}
In this paper we describe and analyze how deoptimization works in the context
of tracing just-in-time compilers. What instructions are used in the
intermediate and low-level representation of the JIT instructions and how these
are implemented.

The goal of this paper is to understand the design constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points. They need to store
enough information to rebuild the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of RPython's
tracing JIT and explain the rationale behind the design decisions based on the
numbers provided here.

\cfbolz{this paragraph now suddenly \emph{introduces} guards, despite having talked about them already}
The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section \ref{sec:Resume Data}). During the recording phase special operations,
referred to as \texttt{guards}, are inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after the
optimization pass over the traced and later compiled parts of the benchmarks,
making guards one of the most common types of operations. Many of these guards
fail rarely or not all during execution. There are several aspects to consider
in the design and optimization of guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the low-level design of a tracing just-in-time compiler.

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind and the implementation of guards in
RPython's tracing just-in-time compiler. the contributions of this paper are:
\begin{itemize}
  \item An analysis of guards in the context of RPython's tracing JIT to
  substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level parts of the JIT and describe the rationale behind the design.
\end{itemize}

\begin{figure}
    \include{figures/guard_table}
    \caption{Percentage of guards before and after optimization for different benchmarks}
    \label{fig:guard_percent}
\end{figure}

The set of central concepts upon which this work is based is described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend\bivab{better term for this?} related to recording and storing the
information required to restore the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section \ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

The RPython language and the PyPy project were started in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative implementation of the Python language\bivab{mention speed}. The
implementation takes advantage of the language features provided by RPython
such as the provided tracing just-in-time compiler described below.

@@

RPython is constructed from two components:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language
is a statically typed object oriented high level language. The language provides
several features such as automatic memory management
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the program to a low level representations suited to be compiled
and run on one of the different supported target platforms/architectures such
as C, .NET and Java. During the transformation process
different low level aspects suited for the target environment are automatically
added to program such as (if needed) a garbage collector and with some hints
provided by the author a just-in-time compiler.

\subsection{RPython's Tracing JIT Compilers}
\label{sub:tracing}
Tracing JITs are a technique of just-in-time compilers that generate code by
observing the execution of a program. VMs using tracing JITs typically are
mixed-mode execution environments that also contain an interpreter. The
interpreter profiles the executing program and selects frequently executed code
paths to be compiled to machine code. Many tracing JIT compilers focus on
selecting hot loops.

After profiling identifies an interesting
path, tracing is started, recording all operations that are executed on this
path. This includes inlining functional calls.
As in most compilers, tracing JITs use an intermediate representation to
store the recorded operations, which is typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}, these operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
In the case of dynamic languages, guards are also used to encode type checks
that come from optimistic type specialization by recording the types of
variables seen during tracing.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

@@

Since tracing linearizes control flow by following one concrete execution,
the full control flow of a program is not observed.
The possible points of deviation from the trace are guard operations
that check whether the same assumptions observed during tracing
still hold during execution.
Similarly, in the case of dynamic languages guards can also encode type
assumptions.
In later executions of the trace the guards can fail.
If that happens, execution needs to continue in the interpreter.
This means it is necessary to attach enough information to a guard
to reconstruct the interpreter state when that guard fails.
This information is called the \emph{resume data}.

@@

After tracing has been finished the trace is optimized.
During optimization a large percentage of operations can be removed.
In the process the resume data is transformed into its final, compressed form.
The rationale for not compressing the resume data during tracing
is that a lot of guards will be optimized away.
For them, the compression effort would be lost.

The core idea of storing resume data as compactly as possible
is to share parts of the data structure between subsequent guards.
This is often useful because the density of guards in traces is so high,
that quite often not much changes between them.
Since resume data is a linked list of symbolic frames
often only the information in the top frame changes from one guard to the next.
The other symbolic frames can often just be reused.
The reason for this is that during tracing only the variables
of the currently executing frame can change.
Therefore if two guards are generated from code in the same function
the resume data of the rest of the frame stack can be reused.

@@

The other important point of interaction between resume data and the optimizer
is RPython's allocation removal optimization~\cite{bolz_allocation_2011}.
This optimization discovers allocations in the trace that create objects
that do not survive long.
An example is the instance of \lstinline{Even} in Figure~\ref{fig:unopt-trace}.
Allocation removal makes resume data more complex.
Since allocations are removed from the trace it becomes necessary
to reconstruct the objects that were not allocated so far when a guard fails.
Therefore the resume data needs to store enough information
to make this reconstruction possible.

Adding this additional information is done as follows.
So far, every variable in the symbolic frames
contains a constant or an SSA variable.
After allocation removal the variables in the symbolic frames can also contain
``virtual'' objects.
These are objects that were not allocated so far,
because the optimizer removed their allocation.
The structure of the heap objects that have to be allocated on guard failure
is described by the virtual objects stored in the symbolic frames.
To this end, the content of every field of the virtual object is described
in the same way that the local variables of symbolic frames are described.
The fields of the virtual objects can therefore be SSA variables, constants
or other virtual objects.
They are encoded using the same compact two-byte representation
as local variables.

@@

Figure~\ref{fig:trace-log} shows the optimized version of the trace in
Figure~\ref{fig:fig:unopt-trace}. Allocation removal has removed the
\lstinline{new} operation and other operations handling the instance. The
operations handle unboxed numbers now.

@@

After optimization the resulting trace is handed to the over platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the to be
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics, operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can often be merged, further reducing even more the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how an \texttt{int\_eq} operation
followed by a guard that checks the result of the operation are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

@@

Each guard in the IR has attached to it a list of the IR-variables required to
rebuild the execution state in case the trace is left through
the guard. When a guard is compiled, in addition to the
condition check two things are generated/compiled. First a special data
structure called \emph{low-level resume data} is created that encodes the
information provided by the register allocator about where the values
corresponding to each IR-variable required by the guard will be stored when
execution reaches the code emitted for the corresponding guard. This data
structure stores the data in a compressed manner using an encoding that uses
8bits to store 7bits of information. This encoding is efficient to create and
provides a compact representation of the needed information,
to maintain an acceptable memory profile.

Second a piece of code is generated for each guard that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline. In case the
condition checked in the guard fails
execution jumps to the corresponding trampoline. In the trampoline the pointer to the
\emph{low-level resume data} is loaded and jumps to generic bailout handler, also known as compensation code,
that is used to leave the compiled trace in case of a guard failure.

Using the encoded location information the bailout handler reads from the
saved execution state the values that the IR-variables had  at the time of the
guard failure and stores them in a location that can be read by the frontend.
After saving the information the control is passed to the frontend signaling
which guard failed so the frontend can read the information passed and restore
the state corresponding to the point in the program.

As in previous sections the underlying idea for the design of guards is to have
a fast on-trace profile and a potentially slow one in the bailout case where
the execution takes one of the side exits due to a guard failure. At the same
time the data stored in the backend needed to rebuild the state needs to be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

As explained in previous sections, when a specific guard has failed often enough
a new trace, referred to as a \emph{bridge}, starting from this guard is recorded and
compiled. When compiling bridges the goal is that future failures of the guards
that led to the compilation of the bridge should execute the bridge without
additional overhead. In particular the failure of the guard should not lead
to leaving the compiled code prior to execution the code of the bridge.

The process of compiling a bridge is very similar to compiling a loop.
Instructions and guards are processed in the same way as described above. The
main difference is the setup phase. When compiling a trace we start with a clean
slate. The compilation of a bridge is started from a state (register and stack
bindings) that corresponds to the state during the compilation of the original
guard. To restore the state needed to compile the bridge we use the encoded
representation created for the guard to rebuild the bindings from IR-variables
to stack locations and registers used in the register allocator.  With this
reconstruction all bindings are restored to the state as they were in the
original loop up to the guard. This means that no register/stack reshuffling is
needed before executing a bridge.

Once the bridge has been compiled the guard that led to compiling the bridge is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard \#1 jumps to the trampoline, loads the \texttt{low level resume data} and
then calls the bailout handler, whereas Guard \#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

\section{Evaluation}
\label{sec:evaluation}
\todo{improve the table formatting}
\todo{give a reference to the benchmark scripts to make things repeatable}

@@

For the different benchmarks the \texttt{low-level resume data} has a size of
about 15\% to 20\% of the amount of memory compared to the size of the
generated machine code. On the other hand the generated machine code has only a
size ranging from 20.5\% to 37.98\% of the size of the high and low-level
resume data combined and being compressed as described before.

Tracing JIT compilers only compile the subset of the code executed in a program
that is traced in a hot loop, for this reason the amount of generated machine
code will be smaller than in other juts-in-time compilation approaches.  This
creates a larger discrepancy between the size of the \texttt{resume data} when
compared to the size of the generated machine code and illustrates why it is important to compress the \texttt{resume data} information.

@@

\subsection{Guard Failures}
\label{sub:guard_failure}
The last point in this discussion is the frequency of guard failures.
Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
relative amounts of guards that ever fail and of guards that fail more than 200
times.\footnote{
    The threshold of 200 is rather high. It was picked experimentally to give
    good results for long-running programs.
}
As described before, for guards that fail more than 200 times, a trace
is recorded that starts from the guard. Afterwards the guard is patched so that later
failures execute the new trace instead of jumping to the trampoline and returning to the interpreter. Hence the
numbers presented for guards that fail more than 200 times represent the 200
failures up to the compilation of the bridge and all executions of the then
attached bridge.

@@

From Figure~\ref{fig:failing_guards} we can see that only a very small amount
of all the guards in the optimized traces ever fail. This amount varies between
2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
enough that a bride is compiled for them, only 1.2\% to 3.6\% of all guards
fail more than 200 times. Also of all failing guards a few fail extremely often
and most fail rarely. The results emphasizes that as most of the guards never
fail it is important to make sure that the successful execution of a guard does
not have unnecessary overhead.

@@

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information
about how to rebuild the state from a guard failure using the information in the
snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
snapshots for guards in LuaJIT are associated with a large memory footprint.
The solution used in there is to store sparse snapshots, avoiding the creation
of snapshots for every guard to reduce memory pressure. Snapshots are only
created for guards after updates to the global state, after control flow points
from the original program and for guards that are likely to fail. As an outlook
Pall mentions the plans to switch to compressed snapshots to further reduce
redundancy. The approach of not creating snapshots at all for every guard is
orthogonal to the resume data compression presented in this paper and could be
reused within RPython to improve the memory usage further.

@@

The deoptimization information of Hotspot is extended
to support correct behaviour
when scalar replacement of fields is done for non-escaping objects~\cite{kotzmann_escape_2005}.
The approach is extremely similar to how RPython's JIT handles virtual objects.
For every object that is not allocated in the code,
the deoptimization information contains a description
of the content of the fields.
When deoptimizing code, these objects are reallocated
and their fields filled with the values
described by the deoptimization information.
The paper does not describe any attempts to store this information compactly.
This may not be needed in their approach, because method-based JITs have
fewer deoptimization points than tracing JITs.

@@

Finally we have presented experimental data collected using the standard PyPy
benchmark set to evaluate previous observations and assumptions. Our
experiments showed that, as previously assumed, guards are a very common
operation in traces. At the same time guards are associated with a high
overhead, because for all compiled guards information needs to be
stored to restore the execution state in case of a bailout. The measurements
showed that the compression techniques used in PyPy effectively reduce the
overhead of guards, while it still produces a significant overhead. The results
also showed that guard failure is a local event: there are few
guards that fail at all, and even fewer that fail very often.
These numbers validate the design decision of reducing the overhead of
successful guard checks as much as possible while paying a higher price in the
case of bailout due to having to decode compressed state representation.
The compressed state representation is reduces the memory footprint of rarely
used data.
